---
title: "Figure 1 - Clinical"
author: "Estef Vazquez"
date: "`r Sys.Date()`"
output: workflowr::wflow_html
editor_options:
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Clinical characteristics of acral melanoma patients according to ulceration status

The following code creates visualizations comparing clinical features between ulcerated and non-ulcerated acral melanomas. It generates box plots for age and mitotic index, violin plot for Breslow depth, stacked bar plots for stage and recurrence, and a combined panel.

```{r libraries}
# Load required libraries
library(tidyverse)
library(patchwork)
library(ggpubr)
library(here)

# Set seed 
set.seed(123)

# Load metadata from 59 patients with complete ulceration status
metadata <- read.csv(here("data", "clinical_am_prim.csv"))

# Convert to factors
metadata <- metadata %>%
  mutate(across(c(-breslow_depth_mm, -mitotic_index, -age, -sample_id, -patient), as.factor))
```

### 1A - Age

```{r, boxplot_age fig.width=12, fig.height=10}
# Boxplot ulceration vs age
ulc_age_plot <- ggplot(data = metadata,
       mapping = aes(x = factor(ulceration), y = age, 
                     fill = factor(ulceration))) + 
  geom_boxplot(na.rm = TRUE,
               width = 0.4,
               alpha = 0.9,
               outlier.shape = NA,
               color = "black"        
               ) + 
  geom_point(size = 2.5, alpha = 0.5, 
             position = position_jitter(seed = 1, width = 0.1),
             color = "black") + 
  stat_summary(fun = mean, 
              geom = "point",
              shape = 23, 
              size = 1.3,
              fill = "white",
              color = "black") +
  stat_compare_means(size = 4, label.x = 1.34) + 
  scale_fill_manual(values=c("#730769", "#E8CC03"),
                    labels = c("No", "Yes")) + 
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
  theme_minimal() +
   theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11),
    legend.position = "right",  
    # Background elements
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  labs(
    x = "Ulceration status", 
    y = "Age at diagnosis",
    title = "Age Distribution by Ulceration Status",
     fill = "Ulceration"
  ) +
  scale_y_continuous(
    limits = function(x) c(floor(min(x)), ceiling(max(x))),
    expand = expansion(mult = c(0.05, 0.05))
  ) + coord_cartesian(xlim = c(0.5, 2.5)) 

ulc_age_plot
```

### 1B - Breslow Depth

```{r, violin_breslow fig.width=12, fig.height=10}
# Violin plot ulceration vs breslow depth 
ulc_bres_plot <- ggplot(metadata, 
            aes(x = factor(ulceration), 
                y = breslow_depth_mm, 
                fill = factor(ulceration))) + 
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1,
               alpha = 0.7,
               color = "black") +
  stat_summary(fun = mean, 
               geom = "point", 
               shape = 23, size = 2,
               color = "black") +
    stat_compare_means(size = 3.9, label.x = 1.36) +
  scale_fill_manual(values = c("#730769", "#E8CC03"),
                    labels = c("No", "Yes")) +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11),
    legend.position = "right",  
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  labs(
    x = "Ulceration status",
    y = "Breslow depth (mm)",
    title = "Ulceration Status and Breslow Depth",
    fill = "Ulceration"
  ) +
  scale_y_continuous(
    limits = function(x) c(floor(min(x)), ceiling(max(x))),
    expand = expansion(mult = c(0.05, 0.05))
  )
ulc_bres_plot
```

### 1C - Mitotic Index

```{r, boxplot_mindex fig.width=12, fig.height=10}
# Boxplot ulceration vs Mitotic Index
ulc_mit_plot <- ggplot(data = metadata,
       mapping = aes(x = factor(ulceration), y = mitotic_index, 
                     fill = factor(ulceration))) + 
  geom_boxplot(na.rm = TRUE,
               width = 0.4,
               alpha = 0.9,
               outlier.shape = NA,
               color = "black",       
               position = position_dodge(width = 0.5)) + 
  geom_point(size = 2.5, alpha = 0.5, 
             position = position_jitter(seed = 1, width = 0.1),
             color = "black") + 
  
  stat_summary(fun = mean, 
              geom = "point",
              shape = 23, 
              size = 1.5,
              fill = "white",
              color = "black") +
  
  stat_compare_means(size = 4,
                       label.x = 1.33) + 
    theme_minimal() +
    theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11),
    legend.position = "right",  
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  scale_fill_manual(values=c("#730769", "#E8CC03"),
                    labels = c("No", "Yes")) + 
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
  labs(
    x = "Ulceration Status", 
    y = "Mitotic Index",
    title = "Mitotic Index by Ulceration Status",
    fill = "Ulceration"
  ) +
  scale_y_continuous(
    limits = function(x) c(floor(min(x)), ceiling(max(x))),
    expand = expansion(mult = c(0.05, 0.05))
  )
ulc_mit_plot
```

### 1D - Tumor Stage

```{r, barplot_stage fig.width=12, fig.height=10}
ulc_stage_plot <- ggplot(metadata, 
       aes(x = stage, 
           fill = factor(ulceration))) +
  geom_bar(position = "stack",
           alpha = 0.8) +
  scale_fill_manual(values = c("#730769", "#E8CC03"),
                    labels = c("No", "Yes")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10, angle = 0),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 11),
    legend.position = "right",
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  ) +
  labs(
    x = "Stage",
    y = "Number of Cases",
    title = "Ulceration Status Across Tumor Stages",
    fill = "Ulceration"
  )
ulc_stage_plot
```

### Tumor recurrence

```{r, fig.width=12, fig.height=10}
ggplot(metadata, 
       aes(x = factor(ulceration), fill = factor(recurrence))) + 
  
  geom_bar(position = "fill",
           width = 0.7,
           alpha = 0.8) +
  geom_text(aes(label = scales::percent(after_stat(count)/sum(after_stat(count)), 
                                      accuracy = 1)),
            stat = "count",
            position = position_fill(vjust = 0.5),
            color = "white",
            size = 4) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#E8CC03", "#730769"),
                    labels = c("No", "Yes")) +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10, color = "black"),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  labs(
    x = "Ulceration Status",
    y = "Proportion of Cases",
    title = "Recurrence in Ulcerated vs Non-ulcerated Tumours",
    subtitle = paste("Total cases:", nrow(metadata)),
    fill = "Recurrence"
  )
```

### Panel

```{r, panel fig.width=12, fig.height=10}

# Combine plots into a single panel
ulc_age_plot <- ulc_age_plot + 
  theme(plot.margin = margin(10, 10, 10, 10, "pt"),
        aspect.ratio = 0.9) +
  coord_cartesian(clip = "off")

ulc_bres_plot <- ulc_bres_plot + 
  theme(plot.margin = margin(10, 10, 10, 10, "pt"),
        aspect.ratio = 0.9) +
  coord_cartesian(clip = "off")

ulc_mit_plot <- ulc_mit_plot + 
  theme(plot.margin = margin(10, 10, 10, 10, "pt"),
        aspect.ratio = 0.9) +
  coord_cartesian(clip = "off")

ulc_stage_plot <- ulc_stage_plot + 
  theme(plot.margin = margin(10, 10, 10, 10, "pt"),
        aspect.ratio = 0.9)


ulc_age_plot <- ulc_age_plot + theme(legend.position = "none")
ulc_bres_plot <- ulc_bres_plot + theme(legend.position = "none")
ulc_mit_plot <- ulc_mit_plot + theme(legend.position = "none")
ulc_stage_plot <- ulc_stage_plot + theme(legend.position = "none")


dummy_plot <- ggplot(metadata, aes(x = factor(ulceration), fill = factor(ulceration))) +
  geom_bar() +
  scale_fill_manual(values = c("#730769", "#E8CC03"),
                   labels = c("No", "Yes"),
                   name = "Ulceration") + theme_minimal()

legend <- cowplot::get_legend(dummy_plot + theme(legend.position = "bottom"))


grid_no_legend <- gridExtra::arrangeGrob(
  ulc_age_plot, ulc_bres_plot,
  ulc_mit_plot, ulc_stage_plot,
  ncol = 2,
  nrow = 2,
  top = ""
)

# Create final grid 
final_grid <- gridExtra::arrangeGrob(
  grid_no_legend, 
  legend,
  ncol = 1,
  heights = c(10, 1)
)

final_grid_with_labels <- cowplot::ggdraw(final_grid) +
  cowplot::draw_grob(final_grid) +
  cowplot::draw_label("a", x = 0.02, y = 0.98, size = 18, fontface = "bold") +
  cowplot::draw_label("b", x = 0.52, y = 0.97, size = 18, fontface = "bold") +
  cowplot::draw_label("c", x = 0.02, y = 0.48, size = 18, fontface = "bold") +
  cowplot::draw_label("d", x = 0.52, y = 0.48, size = 18, fontface = "bold")

final_grid_with_labels

#ggsave("ulceration_combined_panel.pdf", final_grid_with_labels, height = 10, width = 12, dpi = 400)
```

```{r}
sessionInfo()
```
